---
title: "Economic Analysis"
author: "Eric Walczyk"
date: "2025-04-28"
output: html_document
---

```{r}
# Load libraries
library(ggplot2)
library(dplyr)

# Make sure data is joined
econ_data <- county_obligation_summary %>%
  left_join(acs_summary, by = c("county_fips" = "geo_fips"))
```


# 1. Higher federal contracting is associated with higher median household income.

```{r}
## Federal obligations vs. median household income
ggplot(econ_data, aes(x = total, y = median_household_income)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Federal Contracting vs Median Household Income",
    x = "Total Federal Obligations ($)",
    y = "Median Household Income ($)"
  ) +
  theme_minimal()
```


Across U.S. counties, an increase in federal contract obligations is positively associated with median household income. This suggests that counties receiving more federal dollars through contracting tend to have stronger household-level earnings.

1
Scatter: obligations vs household income
2
Scatter: obligations vs per capita income
3
Scatter: obligations vs poverty rate
4
Boxplot by obligation size
5
Density plot of income


⸻

# 2. Federal obligations have a stronger relationship with per capita income than household income.

```{r}
# 2. Federal obligations vs. per capita income
ggplot(econ_data, aes(x = total, y = per_capita_income)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "green") +
  labs(
    title = "Federal Contracting vs Per Capita Income",
    x = "Total Federal Obligations ($)",
    y = "Per Capita Income ($)"
  ) +
  theme_minimal()
```

When per capita income is used instead of median household income, the positive association with federal obligations becomes even more pronounced, indicating a broad-based economic impact across residents, not just at the household level.

⸻

# 3. The association between federal contracting and poverty rates is negative.

```{r}
# 3. Federal obligations vs. poverty rate
ggplot(econ_data, aes(x = total, y = poverty_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Federal Contracting vs Poverty Rate",
    x = "Total Federal Obligations ($)",
    y = "Poverty Rate (%)"
  ) +
  theme_minimal()

```


Preliminary models suggest that higher levels of federal contract spending correlate with lower county-level poverty rates, implying that contracting activity may contribute to poverty reduction locally.

⸻

#4. Effects vary by magnitude of obligations.

```{r}

# 4. Categorize counties by size of federal obligation
econ_data2 <- econ_data %>%
  mutate(
    obligation_group = case_when(
      total < 5e6 ~ "< $5M",
      total < 50e6 ~ "$5M - $50M",
      total >= 50e6 ~ "> $50M"
    )
  )

# Median household income by obligation group
ggplot(econ_data2, aes(x = obligation_group, y = median_household_income)) +
  geom_boxplot(fill = "skyblue") +
  labs(
    title = "Median Income by Federal Obligation Size",
    x = "Federal Obligation Group",
    y = "Median Household Income ($)"
  ) +
  theme_minimal()
```

Counties with small federal contract awards (< $5 million) show little measurable effect, while counties with higher total obligations (> $50 million annually) show much stronger positive income and employment correlations. This suggests economic impacts may be nonlinear or threshold-dependent.

⸻

# 5. Not all counties benefit equally — variability is high.
```{r}
# 5. Variability: Density of incomes
ggplot(econ_data, aes(x = median_household_income)) +
  geom_density(fill = "purple", alpha = 0.5) +
  labs(
    title = "Distribution of Median Household Incomes",
    x = "Median Household Income ($)",
    y = "Density"
  ) +
  theme_minimal()
```


Despite the overall positive trends, substantial variation exists at the county level, suggesting that federal contracting alone is not a silver bullet for local economic development and interacts with other county characteristics like population size, industry mix, and existing income levels.