---
title: "Untitled"
author: "Eric Walczyk"
date: "2025-04-06"
output: html_document
---

## recleaning the federal contract data because i forgot to add filters
## for small businesses. Will also add filters for other business categories
## (woman-owned, minority-owned, veteran-owned, etc.)

## Because there is 109 GB of data, I will query and compile it through 
## by connecting to my DBeaver database using a .duckdb file. This lets me
## filter and clean the entire dataset without having to actually load them
## all into Rstudio.

```{r}
con <- dbConnect(duckdb::duckdb(), dbdir = "fedcon.duckdb", read_only = FALSE)
```

```{r}
process_fiscal_year <- function(fy) {
  # Construct table and file names based on fiscal year
  contracts_table <- sprintf("fy%s_contracts", fy)
  summary_table   <- sprintf("fy%s_summary", fy)
  csv_path        <- sprintf("/Volumes/vorb/02_school/capstone/SBA-Capstone-Dashboard/data/raw/usa_spending/FY%s/*_clean.csv", fy)
  output_path     <- sprintf("/Volumes/vorb/02_school/capstone/SBA-Capstone-Dashboard/data/cleaned/fy%s_summary.csv", fy)
  
  # Connect to DuckDB
  con <- dbConnect(duckdb::duckdb(),
                   dbdir = "/Volumes/vorb/02_school/capstone/SBA-Capstone-Dashboard/data/fedcon.duckdb")

  # Load all CSV chunks from the BOM-cleaned files into a contracts table
  query_contracts <- sprintf("
    CREATE OR REPLACE TABLE %s AS
    SELECT * FROM read_csv_auto('%s', ignore_errors=true)
  ", contracts_table, csv_path)
  dbExecute(con, query_contracts)
  
  # Summarize data: group by the needed columns and sum federal_action_obligation
  query_summary <- sprintf("
    CREATE OR REPLACE TABLE %s AS
    SELECT
      %s AS fiscal_year,
      prime_award_transaction_place_of_performance_county_fips_code AS county_fips,
      primary_place_of_performance_state_code AS state,
      awarding_agency_name AS agency,
      naics_code,
      naics_description,
      
      -- Flags
      UPPER(contracting_officers_determination_of_business_size) = 'SMALL BUSINESS' AS is_small_business,
      woman_owned_business = 'Y' AS is_woman_owned,
      veteran_owned_business = 'Y' AS is_veteran_owned,
      minority_owned_business = 'Y' AS is_minority_owned,

      -- Total obligation for group
      SUM(federal_action_obligation) AS total_obligation

    FROM %s
    WHERE 
      prime_award_transaction_place_of_performance_county_fips_code IS NOT NULL
      AND federal_action_obligation IS NOT NULL
      AND federal_action_obligation >= 0
      AND UPPER(contracting_officers_determination_of_business_size) = 'SMALL BUSINESS'

    GROUP BY 
      county_fips, state, agency, naics_code, naics_description,
      is_small_business, is_woman_owned, is_veteran_owned, is_minority_owned
  ", summary_table, fy, contracts_table)
  dbExecute(con, query_summary)

  # Export summary to CSV
  query_export <- sprintf("
    COPY %s TO '%s' (HEADER, DELIMITER ',')
  ", summary_table, output_path)
  dbExecute(con, query_export)

  # Disconnect from the database
  dbDisconnect(con)
  
  # Message
  cat("Fiscal Year", fy, "processed and exported to", output_path, "\n")
}

# Process fiscal years 2019 through 2025
for(fy in 2017:2025) {
  process_fiscal_year(fy)
}
```

```{r}
  # Disconnect from the database
  dbDisconnect(con)
```

